---
title: "SC306 Homework 4"
author: "Othar Zaldastani II"
date: "March 9, 2026"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{xcolor}
- \makeatletter
- \makeatother
- \usepackage{float}
- \usepackage{endnotes}
- \renewcommand{\CommentTok}[1]{\textcolor[rgb]{0.60,0.20,0.20}{#1}}
---

```{r setup, echo=TRUE, results='hide', message=FALSE, warning=FALSE, include = F}
knitr::opts_chunk$set(fig.pos = "H", message = FALSE, warning = FALSE)
library(tidyverse)
library(dplyr)
library(epitools)
library(knitr)
```

## Question 1

Creation of birth weight table

```{r}
bw = matrix(c(110, 1073, 289, 15896), nrow=2, byrow = TRUE)
colnames(bw) = c('Low', 'Normal')
rownames(bw) = c('Cocaine', 'No Cocaine')
knitr::kable(bw)
```

#### A)

The incidence measure we can calculate given the data is incidence proportion. $IP = \frac {\text{Total number of new cases during interval}} {\text{Total population at risk}}$.

#### B)

$IP_{Cocaine} = \frac{110}{110+289} = 0.276$

#### C)

$IP_{Non-Cocaine} = \frac{1073}{1073+15896} = 0.0063$

#### D)

The relative risk of low birth weight is $\frac{0.276}{0.063} = 4.38$.

#### E)

A relative risk of 4.38 means mothers who used Cocaine during pregnancy had 4.38 times the risk of having a baby with low birth weight compared to mothers who did not use Cocaine during pregnancy.

#### F)

Birth weight table

```{r}
riskratio(bw, rev = "b")
```

The risk ratio is 5.21 with a 95% CI of $[4.21 : 6.43]$. The interpretation for the 95% confidence interval is that given our analysis of the data, we are 95% confident that the true risk ratio is between 4.21 and 6.43. Cocaine use is significantly associated with low birth weight, as shown by the value lower bound of our confidence interval considerably exceeding 1.

## Question 2

```{r}
IPR.conf = function(a, b, c, d, conf) {
  ipr.hat = (a/(a+b)) / (c/(c+d))                 # Predicted IPR
  alpha = 1 - conf
  z = qnorm(1-alpha/2)
  
  var.ln.ipr = (b/(a*(a+b))) + (d/(c*(c+d)))      # Natural log of IPR variance
  se = sqrt(var.ln.ipr)                           # Standard error
  
  ln.lower = log(ipr.hat) - z * sqrt(var.ln.ipr)  # Upper/Lower Log Bounds
  ln.upper = log(ipr.hat) + z * sqrt(var.ln.ipr)
  
  lower = exp(ln.lower)                           # Upper/Lower Bounds
  upper = exp(ln.upper)
  
  data.frame(
    IPR = ipr.hat,
    "Lower Bound" = lower,
    "Upper Bound" = upper,
    "Confidence Interval" = conf,
    check.names = F
  )}
kable(IPR.conf(110, 1073, 289, 15896, .95), digits = 3)
```

The new interval is effectively identical to the interval from Problem 1.

## Question 3: BMI and Mortality

#### A)

The exposure factor is BMI category.

#### B)

The outcome is all-cause mortality.

#### C)

Incidence density can be computed using this data because the follow-up is measured in person-years.

#### D)

Completed table

```{r echo=FALSE}
bmi <- c("Normal","Underweight","Overweight","All")
py <- c(5995,8479,1562,16036)
deaths <- c(19,71,7,97)
incidence <- c(0.00317,0.00837,0.00448,0.00605)
rr <- c(1.00,2.64,1.41,NA)

table <- data.frame(bmi, py, deaths, incidence, rr)

kable(
  table,
  col.names = c(
    "BMI",
    "Follow up (person-years)",
    "Number of deaths",
    "Incidence of mortality",
    "Relative Risk"
  ),
  align = "lcccc"
)
```

#### E)

The underweight population are at greater risk of mortality than the overweight population. The underweight population has an all cause mortality rate 2.64 times higher than normal, while the overweight population has an all cause mortality rate 1.41 times higher than normal.

## Question 4: Clinical Epidemiology

#### A)

Specificity = $\frac{460}{460+40} = 0.92$

#### B)

Sensitivity = $\frac{840}{840+160} = 0.84$

#### C)

The positive predictive value of this test is: $\frac{840}{840+40} = 0.955$

#### D)

The negative predictive value of this test is $\frac{460}{460+160} = 0.742$

## Question 5

#### A)

Because this test is not inflated by a disproportionate number of responses in one category, we use the percent agreement formula: $\frac{A+D}{A+B+C+D}$.

Percent agreement here is $\frac{110+61}{110+30+5+61} = 0.830$

#### B)

$\kappa = \frac{\text{\% agreement - \% agreement due to chance}}{\text{100\% - \% agreement due to chance}}$\newline
Expected in High, High: $\left(\frac{91}{206}\Big)\Big(\frac{66}{206}\right)\times 206 = 29.2$ \newline
Expected in Low, Low: $\left(\frac{115}{206}\right)\left(\frac{140}{206}\right)\times206 = 78.2$ \newline
Expected agreement proportion: $\frac{29.2 + 78.2}{206} = 0.521$ \newline
Observed agreement: 0.830 (from before). \newline
Then, $\kappa = \frac{0.830 - 0.521}{1-0.521} = 0.645$.



## Question 6
Total population: $303 + 162 + 354 + 290 = 1109$


#### A)
The prevalence of BCC is $\frac{303 + 354}{1109} = 0.592$ or 59.2%

#### B)
No, because this is a case-control study, so a non-representative population exists because people with BCC are intentionally selected for the study.

#### C)
The odds ratio is the odds of exposure among cases divided by the odds of exposure of the controls. \newline
$OR = \frac{303/354}{162/290} = 1.53$

#### D)
The odds ratio means the odds of tanning bed use among those with Basal Cell Carcinoma are 1.53 times higher than those without Basal Cell Carcinoma.

#### E)
$OR_{disease} = \frac {303/162}{354/290} = 1.53$
The odds ratio given disease is 1.53.

#### F)
They are the exact same value. The odds ratio is the same if you calculate odds of disease given exposure vs. odds of exposure given disease.

#### G)
No, since this is a case-control study, the number of cases and controls observed was a deliberate choice by those conducting the study, not numbers representative of the general population.