---
title: "SC306 Homework 4"
author: "Othar Zaldastani II"
date: "March 9, 2026"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{xcolor}
- \makeatletter
- \makeatother
- \usepackage{float}
- \usepackage{endnotes}
- \renewcommand{\CommentTok}[1]{\textcolor[rgb]{0.60,0.20,0.20}{#1}}
---

```{r setup, echo=TRUE, results='hide', message=FALSE, warning=FALSE, include = F}
knitr::opts_chunk$set(fig.pos = "H", message = FALSE, warning = FALSE)
library(tidyverse)
library(dplyr)
library(epitools)
library(knitr)
```

## Question 1
Creation of birth weight table
```{r}
bw = matrix(c(110, 1073, 289, 15896), nrow=2, byrow = TRUE)
colnames(bw) = c('Low', 'Normal')
rownames(bw) = c('Cocaine', 'No Cocaine')
knitr::kable(bw)
```
#### A)
The incidence measure we can calculate given the data is incidence proportion.
$IP = \frac {\text{Total number of new cases during interval}} {\text{Total population at risk}}$.

#### B)
$IP_{Cocaine} = \frac{110}{110+289} = 0.276$

#### C)
$IP_{Non-Cocaine} = \frac{1073}{1073+15896} = 0.0063$

#### D)
The relative risk of low birth weight is $\frac{0.276}{0.063} = 4.38$.

#### E)
A relative risk of 4.38 means mothers who used Cocaine during pregnancy had 4.38 times the risk of having a baby with low birth weight compared to mothers who did not use Cocaine during pregnancy.

#### F)
Birth weight table
```{r}
riskratio(bw, rev = "b")
```
The risk ratio is 5.21 with a 95% CI of $[4.21 : 6.43]$. The interpretation for the 95% confidence interval is that given our analysis of the data, we are 95% confident that the true risk ratio is between 4.21 and 6.43. Cocaine use is significantly associated with low birth weight, as shown by the value lower bound of our confidence interval considerably exceeding 1.


## Question 2
```{r}
IPR.conf = function(a, b, c, d, conf) {
  ipr.hat = (a/(a+b)) / (c/(c+d))                 # Predicted IPR
  alpha = 1 - conf
  z = qnorm(1-alpha/2)
  
  var.ln.ipr = (b/(a*(a+b))) + (d/(c*(c+d)))      # Natural log of IPR variance
  se = sqrt(var.ln.ipr)                           # Standard error
  
  ln.lower = log(ipr.hat) - z * sqrt(var.ln.ipr)  # Upper/Lower Log Bounds
  ln.upper = log(ipr.hat) + z * sqrt(var.ln.ipr)
  
  lower = exp(ln.lower)                           # Upper/Lower Bounds
  upper = exp(ln.upper)
  
  data.frame(
    IPR = ipr.hat,
    "Lower Bound" = lower,
    "Upper Bound" = upper,
    "Confidence Interval" = conf,
    check.names = F
  )}
kable(IPR.conf(110, 1073, 289, 15896, .95), digits = 3)
```
The new interval is effectively identical to the interval from Problem 1.




