---
title: "Zaldastani_Othar_cc5"
output:
  pdf_document: default
  html_document: default
---

```{r setup, echo=TRUE, results='hide', message=FALSE, warning=FALSE, include = T}
knitr::opts_chunk$set(fig.pos = "H", message = FALSE, warning = FALSE)
library(tidyverse)
library(modeldata)
data(Sacramento)
library(ggplot2)
```



## Question 1: Visualizing the Relationship

#### A)
Square Footage vs. Price
```{r}
ggplot(data = Sacramento, aes(x = sqft, y = price)) +
  geom_point() +
  xlab("Square Footage") +
  ylab("Price") +
  ggtitle("Housing Price vs. Square Footage") +
  theme(plot.title = element_text(hjust = 0.5))
```

#### B)

Square Footage vs. Price with line of best fit
```{r}
ggplot(data = Sacramento, aes(x = sqft, y = price)) +
  geom_point() +
  xlab("Square Footage") +
  ylab("Price") +
  ggtitle("Housing Price vs. Square Footage") +
  theme(plot.title = element_text(hjust = 0.5))+
  geom_smooth(method = "lm")
```

#### C)
I expect $\hat \beta_1$ to be positive because the line of best fit has a positive slope.


## Question 2: From Bivariate to Multivariate

#### A)

```{r}
biv.model = lm(price ~ sqft, data = Sacramento)
summary(biv.model)
```
The coefficient on `sqft` is 138.55.

#### B)

```{r}
multi.model = lm(price ~ sqft + beds, data = Sacramento)
summary(multi.model)
```

The coefficient on `sqft` is now 161.34 and the coefficient on `beds` is -26035.15.

#### C)
The coefficient on `sqft` increased by over 16% after adding `beds` to the regression.

#### D)
The coefficient on `sqft` likely changed after adding `beds` as a predictor variable because when holding square footage fixed, more bedrooms would imply smaller bedrooms and less space per person in a household. A 4 bedroom, 1500 square foot house is likely more cramped than a house of the same size with two bedrooms.


## Question 3: Interpreting Coefficients and Adding Controls

#### A)
Adding `baths` to the model.
```{r}
A3.model = lm(price ~ sqft + beds + baths, data = Sacramento)
summary(A3.model)
```

#### B)

Holding bedrooms and bathrooms constant, 1 square foot size increase raises a house's predicted sale price by $158.17.

#### C)

When holding bedrooms and square footage constant, an an increase in bathroom count of 1 raises its predicted price by $5342.59. However, it is worth noting that when holding `sqft` and `beds` constant, `baths` is not a statistically significant predictor of a house's market value.

#### D)
Tibble creation
```{r}
comparison = tibble(
  model = c("biv.model",
            "multi.model",
            "A3.model"),
  sqft_coef = c(coef(biv.model)["sqft"],
                coef(multi.model)["sqft"],
                coef(A3.model)["sqft"]))
comparison
```

#### E)

Adding `baths` did not substantially change the the coefficient on `sqft` compared to the model with only `sqft` and `beds`. This suggests `baths` is not an important confounding variable for the square footage–price relationship.


## Question 4: Omitted Variable Bias
Data Generation
```{r}
set.seed(325)
n <- 500

# True DGP: y = 10 + 3*x1 + 5*x2 + error
# x1 and x2 are correlated [cor(x1, x2) = 0.6]
x2 <- rnorm(n)
x1 <- 0.6 * x2 + rnorm(n)
y <- 10 + 3 * x1 + 5 * x2 + rnorm(n)

sim_data <- tibble(y, x1, x2)
```


#### A)
The true value of $\beta_1$ is 3.\

#### B)





