---
title: "SC306 Homework 5"
author: "Othar Zaldastani II"
date: "March 17, 2026"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{xcolor}
- \makeatletter
- \makeatother
- \usepackage{float}
- \renewcommand{\CommentTok}[1]{\textcolor[rgb]{0.60,0.20,0.20}{#1}}
---

```{r setup, echo=TRUE, results='hide', message=FALSE, warning=FALSE, include = F}
knitr::opts_chunk$set(
  echo = FALSE,                 # don't show code
  warning = FALSE,              # don't show warnings
  message = FALSE,              # don't show messages (less serious warnings)
  cache = FALSE,                # set to TRUE to save results from last compilation
  fig.align = "center",         # center figures
  attr.source = '.numberLines'  # add line numbers to code
)
library(tidyverse)
library(dplyr)
library(knitr)
library(epitools)

set.seed(24)

```


# Question 1

#### A) 
$AR \ \% = \frac{R_e - R_u}{R_e} \times 100$


```{r}
# Question 1
# Part A
Re = (110)/(110+289) # Risk among exposed

Ru = (1073)/(1073+15896) # Risk among unexposed

AR = (Re - Ru)/(Re) * 100 # Attributable risk

round(AR, 1)
```

The attributable risk percent of low birth weight from Cocaine use is 77.1%.

#### B)
This means approximately 77% of the low birth weight risk for infants born to mothers who used Cocaine while pregnant is attributable to their use of Cocaine.

#### C) 
$PAR \ \% = \frac{P_e(R_e - R_u)}{P_e R_e + (1-P_e)R_u} \times 100$
```{r}
# Part C
Pe = 0.02
PAR = (Pe * (Re - Ru)) / (Pe * Re + (1-Pe) * Ru) * 100
round(PAR, 2)
```
The population attributable risk percent of low birth weight from Cocaine use is 6.3%.

#### D) 
Approximately 6.3% of the low birth weight risk for all infants born in California is attributable to the use of Cocaine while pregnant.




\pagebreak
# Code Appendix
```{r ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```

\vfill

Attributions: I have used some template/styling code from GitHub user alexpghayes.